Synthesizing Symmetric Lenses

### 2018-06-29 papers/SymmetricLenses/main.tex
Comments and a bit of polishing

Lens languages, like any domain-specific language, also cut down the space of
possible programs one can write, and the smaller the space of programs,
the easier it is to find the program one is looking for.  
%This feature can be a negative, if the desired transformation
%falls outside of the scope provided by the language, but can also be positive, as complex
%properties (like invertability) can be enforced automatically.
%Another advantage of reducing the
%set of programs is that doing so can facilitate program synthesis, which is essentially
%a search through the space of possible computations that can be expressed as a lens.
Recent research~\cite{optician}
has exploited this fact by demonstrating it is possible to synthesize relatively
large, complex lenses from specifications.
In particular, given types that describe source and target formats, there are synthesis
algorithms that will find bijective lenses~\cite{optician}
for medium-sized files, such as those housing linux
application configurations.  \ksf{why medium sized?  You can have very
large files with simple format descriptions that are bijective and for
which our previous systems would have worked just fine.} Oftentimes, these algorithms find such lenses in less than
a second, and without the aid of any examples at all, with just types describing the source and
target formats.  \dpw{I would cite the ICFP paper (cite string: maina+:quotient-synthesis) as to appear but I'm not sure how to do that
without possibly violating the spirit of double-blind reviewing. ie, if it hasn't appeared yet,
how could we know the content if we aren't the authors?  Mentioning it in the 3rd person does not
seem to suffice ...} \ksf{The ICFP 2018 website lists the paper; I
think it is okay to cite it.  The reviewer can assume we contacted the
authors for a copy.  We could ask Stephanie for a ruling if we want to
be sure.  One of the principles of double blind as adopted by SIGPLAN
is not to reduce the quality of the paper; omitting this citation
woudl have that affect.}

@@

In order to characterize the expressive power of simple symmetric lenses
theoretically, we prove simple symmetric lenses
include all (and only) those symmetric lenses that satisfy
an intuitive property called \emph{forgetfulness}.  In other words, they are the
complete set of symmetric lenses that do not require a ``memory'' to retain data
from past instances of $S$ or $T$ when making a round trip.\dpw{fix that to make the intuitive
explaination sufficiently accurate}
From a pragmatic perspective, we justify our design choices by extending
the Boomerang implementation with symmetric lenses (they had never before been implemented),
applying them successfully to a range of real-world applications,
and demonstrating that they are compatible, not just with basic bijective combinators,
but with a rich set of advanced features including quotient lenses and matching lenses.

The second difficulty appears to be more fundamental\dpw{better word?}:
While the number of bijective lenses
between two related formats is often tiny, the number of symmetric lenses is often
large.  In practice, the difference in cardinality means that when a search algorithm selects the smallest
bijective lens it finds, that lens has a good chance of being the
one a user wants.  In contrast, and to our surprise, our experiments show \dpw{can we have an experiment that
shows this?} that
when a search algorithm selects the smallest symmetric lens, it typically fails
to find the lens the user wants.  The flexibity these lenses admit when it comes to
eliding information generates an explosion of possible transformations.
To disambiguate between the many possible lenses, we need
new principles for identifying the ``likely'' ones, and we need
new algorithms to search the space of program transformations using these principles.
\dpw{Were the bijective lenses chosen using a ``smaller'' criteria?  I
forget... if not edit the prior paragraph.}\dpw{Ideally, we could back up the claim
that there are many symmetric lenses and not many bijective lenses by graphing the
number of such lenses somehow.  eg, in a past synthesis paper with Steve (POPL 2016)
we graphed
the untyped terms of type $\mathsf{int}\rightarrow\mathsf{int}$.  I wonder if we can
do something similar here.}

To overcome this second difficulty, we turned to information theory
to try to find a way to characterize the likely and unlikely lenses.  Intuitively, a ``likely'' lens
is one that propagates a lot of information from source to target and also from target
back to source.  Conversely, given the source, the target is ``unlikely'' to require very much
additional information in order to reconstruct it---if it were, then one is unlikely to
be attempting to synthesize a lens between the two formats!  In order to implement
this intuition, we turn to \emph{stochastic regular expressions} (SREs).
Stochastic regular expressions define a set of strings, as do ordinary regular expressions,
but they also provide the expected distribution of strings within that set.  Such
distributions allow us to compute the amount of information gained when presented
an element of the set.  For example, the language of the SRE $a |_{0.5} b$ contains two elements
(the strings $a$ and $b$).  The probability 0.5 annotating the union indicates that both
elements are equally likely to appear.  Consequently, if we observe 
an element of that language (either $a$ or $b$), we will gain one bit of information.  Said another way,
one bit, together with the SRE, suffices to recover the string in question.
On the other hand, consider the language of the SRE $c$.
It contains just one element (the string $c$).  If we observe that our algorithm
generates $c$, we gain zero bits of information --- the string was already determined.
Hence, a lens is ``likely,'' in a technical sense, if relative to the target SRE $T$,
very little information from $t$ is required to generate it.

With SREs in hand, we develop a new algorithm for synthesizing likely lenses.
This algorithm is constructed from a pair of collaborating search procedures.
Given source and target SREs $S$ and $T$, the first algorithm searches for two \emph{related} SREs
$S'$ and $T'$ that are \emph{compatible} (a heuristic that estimates the likelihood
our second algorithm will succeed).  $S'$ and $T'$ are related in the sense
that they define the same regular sets as $S$ and $T$ (respectively) and the distribution
of strings is identical.  Indeed, we guarantee these two properties by implementing
a search procedure that obeys the star semi-ring axioms of regular expression equivalence,
and which we have proven does not alter the probability distributions of the languages.

Given the two compatible SREs,
the second algorithm uses a greedy, type-directed algorithm find a likely lens between them.
If algorithm estimates the lens it has discovered is more likely than
any future lens it can possibly find, the search will be terminated.  Otherwise,
this lens is stored away (if it is the best found so far) and the search continues with
algorithm 1 again.

We implemented this algorithm and evaluated its effectiveness on a range of benchmarks
from XXXX to YYYYY. We found that ... \dpw{give a number of two that characterize
our empirical findings.}

In summary, our central contributions are as follows:
\begin{itemize}
\item We define the space of simple symmetric lenses, which lies between asymmetric lenses and full symmetric lenses and has the useful property that they don't have any internal state.  We characterize
the simple lenses semantically as the subset of full symmetric lenses for which a ``forgetfulness''
property holds. We also analyze the gap between simple and full symmetric lenses empirically and argue that
simple ones suffice for many applications.
\item We show how to synthesize simple symmetric lenses (including all asymmetric lenses as a special case),
using a novel application of stochastic regular expressions to guide the search process.  In particular, we
show that the RE star semi-ring equivalences remain valid when their semantics are enriched by distributions.
\item We extend the Boomerang implementation with simple symmetric lenses, showing that they can easily be
combined with a rich variety of other lens extensions including quotients and matching lenses.  We evaluate
our implementation over a set of benchmarks drawn from XXXXX and demonstrate that YYYY ... numbers ....
\end{itemize}

>>>>

Lens languages, like any domain-specific language, also cut down the space of
possible programs one can write, and the smaller the space of programs,
the easier it is to find the program one is looking for.  
%This feature can be a negative, if the desired transformation
%falls outside of the scope provided by the language, but can also be positive, as complex
%properties (like invertability) can be enforced automatically.
%Another advantage of reducing the
%set of programs is that doing so can facilitate program synthesis, which is essentially
%a search through the space of possible computations that can be expressed as a lens.
Recent research~\cite{optician}
has exploited this fact by demonstrating it is possible to synthesize relatively
large, complex lenses from specifications.
In particular, given types that describe source and target formats, there are synthesis
algorithms that will find bijective lenses~\cite{optician}
for medium-sized files, such as those housing linux
application configurations.  \ksf{why medium sized?  You can have very
large files with simple format descriptions that are bijective and for
which our previous systems would have worked just fine.} Oftentimes, these algorithms find such lenses in less than
a second, and without the aid of any examples at all, with just types describing the source and
target formats.  \dpw{I would cite the ICFP paper (cite string: maina+:quotient-synthesis) as to appear but I'm not sure how to do that
without possibly violating the spirit of double-blind reviewing. ie, if it hasn't appeared yet,
how could we know the content if we aren't the authors?  Mentioning it in the 3rd person does not
seem to suffice ...} \ksf{The ICFP 2018 website lists the paper; I
think it is okay to cite it.  The reviewer can assume we contacted the
authors for a copy.  We could ask Stephanie for a ruling if we want to
be sure.  One of the principles of double blind as adopted by SIGPLAN
is not to reduce the quality of the paper; omitting this citation
woudl have that effect.}\bcp{+1}

@@

In order to characterize the expressive power of simple symmetric lenses
theoretically, we prove simple symmetric lenses
include all (and only) those symmetric lenses that satisfy
an intuitive property called \emph{forgetfulness}.  In other words, they are the
complete set of symmetric lenses that do not require a ``memory'' to retain data
from past instances of $S$ or $T$ when making a round trip.\dpw{fix that to make the intuitive
explaination sufficiently accurate}
From a pragmatic perspective, we justify our design choices by extending
the Boomerang implementation~\cite{Boomerang} with symmetric lenses (which
had never before been implemented\bcp{Confusing: we implement only simple
  ones, but it was the earlier full form that had never been implemented}),
applying them successfully to a range of real-world applications,
and demonstrating that they are compatible, not just with basic bijective
combinators,\bcp{confusing: they are not exactly ``compatible'' with the
  bijective combinators---they enrich them}
but with a rich set of advanced features including quotient lenses and matching lenses.

The second difficulty with synthesizing symmetric lenses is more fundamental\dpw{better word?}:
While the number of bijective lenses
between two related formats is often tiny, the number of symmetric lenses is often
large.  In practice, the difference in cardinality means that when a search algorithm selects the smallest
bijective lens it finds, that lens has a good chance of being the
one a user wants.  In contrast, and to our surprise, our experiments show \dpw{can we have an experiment that
shows this?} that
when a search algorithm selects the smallest symmetric lens, it typically fails
to find the lens the user wants.  The flexibity these lenses admit when it comes to
eliding information generates an explosion of possible transformations.
To disambiguate between the many possible lenses, we need
new principles for identifying the ``likely'' ones, and we need
new algorithms to search the space of program transformations using these principles.
\dpw{Were the bijective lenses chosen using a ``smaller'' criteria?  I
forget... if not edit the prior paragraph.}\dpw{Ideally, we could back up the claim
that there are many symmetric lenses and not many bijective lenses by graphing the
number of such lenses somehow.  eg, in a past synthesis paper with Steve (POPL 2016)
we graphed
the untyped terms of type $\mathsf{int}\rightarrow\mathsf{int}$.  I wonder if we can
do something similar here.}

To overcome this second difficulty, we turned to information theory
to try to find a way to characterize the likely and unlikely lenses.  Intuitively, a ``likely'' lens
is one that propagates a lot of information from source to target and also from target
back to source.  Conversely, given the source, the target is ``unlikely'' to require very much
additional information in order to reconstruct it---if it were\bcp{were
  what?  likely, or unlikely?}, then one is unlikely to
be attempting to synthesize a lens between the two formats\bcp{why? even if
  two big formats only share a little bit of common information, mightn't
  one want a lens between them?}!  In order to implement
this intuition, we turn to \emph{stochastic regular expressions} (SREs).
Stochastic regular expressions define a set of strings, as do ordinary regular expressions,
but they also provide the expected distribution of strings within that set.  Such
distributions allow us to compute the amount of information gained when presented
an element of the set.  For example, the language of the SRE $a |_{0.5} b$ contains two elements
(the strings $a$ and $b$).  The probability 0.5 annotating the union indicates that both
elements are equally likely to appear.  Consequently, if we observe 
an element of that language (either $a$ or $b$), we will gain one bit of information.  Said another way,
one bit, together with the SRE, suffices to recover the string in
question.\bcp{Is this implying that, if the annotation were
  different, we'd need something other than one bit to recover the string??}
On the other hand, consider the language of the SRE $c$.
It contains just one element (the string $c$).  If we observe that our algorithm
generates $c$, we gain zero bits of information --- the string was already determined.
Hence\bcp{?}, a lens is ``likely,'' in a technical sense, if relative to the target SRE $T$,
very little information from $t$ is required to generate it.

With simple symmetric lenses and SREs in hand, we develop a new algorithm for synthesizing likely lenses.
This algorithm is constructed from a pair of collaborating search procedures.
Given source and target SREs $S$ and $T$, the first algorithm searches for two \emph{related} SREs
$S'$ and $T'$ that are \emph{compatible} (a heuristic that estimates the likelihood
our second algorithm will succeed).  $S'$ and $T'$ are related in the sense
that they define the same regular sets as $S$ and $T$ (respectively) and the distribution
of strings is identical.  Indeed, we guarantee these two properties by implementing
a search procedure that obeys the star semi-ring axioms of regular expression equivalence,
and which we show does not alter the probability distributions of the languages.

Given the two compatible SREs,
the second algorithm uses a greedy, type-directed algorithm find a likely lens between them.
If algorithm estimates that the lens it has discovered is more likely than
any future lens it can possibly find, the search terminates.  Otherwise,
this lens is stored away (if it is the best found so far) and the search continues with
algorithm 1 again.

We implemented this\bcp{which?} algorithm and evaluated its effectiveness on a range of benchmarks
from XXXX to YYYYY. We found that ... \dpw{give a number or two that characterize
our empirical findings.}

In summary, our main contributions are:\bcp{annotate with section numbers}
\begin{itemize}
\item We define the space of simple symmetric lenses, which lies between asymmetric lenses and full symmetric lenses and has the useful property that they don't have any internal state.  We characterize
the simple lenses semantically as the subset of full symmetric lenses for which a ``forgetfulness''
property holds. We also analyze the gap between simple and full symmetric lenses empirically and argue that
simple ones suffice for many applications.
\item We show how to synthesize simple symmetric lenses (including all asymmetric lenses as a special case),
using a novel application of stochastic regular expressions to guide the search process.  In particular, we
show that the RE star semi-ring equivalences remain valid when their semantics are enriched by distributions.
\item We extend the Boomerang implementation with simple symmetric lenses, showing that they can easily be
combined with a rich variety of other lens extensions including quotients and matching lenses.  We evaluate
our implementation over a set of benchmarks drawn from XXXXX and demonstrate that YYYY ... numbers ....
\end{itemize}
